---
title: "Data Management"
format: 
  pdf: 
    number-sections: true 
    toc: true            
    pdf-engine: xelatex
    geometry: margin=1in
    fontsize: 11pt
editor: visual
---

```{r library}
library(tidyverse)
library(descr)
library(knitr)
library(dplyr)
library(readr)
library(readxl)
```

===My Data===

```{r load-data}
file_path <- "/cloud/project/data/data.xls"

data.xls <- read_excel(file_path)
options(width=80)
head(data.xls)
```

===Frequencys===

```{r}
freq(as.ordered(data.xls$A3.8), plot = FALSE)
```

===Frequency of Religious Orientations===

```{r}
freq(as.ordered(data.xls$A3.8), plot = FALSE)

```

===Frequency of Levels of Educations===

```{r education levels}
freq(as.ordered(data.xls$A3.10), plot = FALSE)

```

This frequency table shows the levels of education within a sample. It shows me how many people don't have a high school degree to how many people have a doctorates. This will help me relate the info to people who are religious,to their levels of education. I need this info because I will eventually compare the information I gather to how many took COVID-19 as a serious threat.

===Frequency of who believed COVID-19 was exaggerated===

```{r COVID-19 exaggerated}
freq(as.ordered(data.xls$`A5.3*`), plot = FALSE)

```

This frequency table shows the levels in which how many people thought COVID-19 was being exaggerated. It goes from level 1 = Not at all to level 10= Extremely. This will help me see how seriously people took COVID-19 and who underestimated it.

===Frequency in which mask was worn===

```{r mask frequency}
freq(as.ordered(data.xls$A31.4), plot = FALSE)

```

This frequency table shows how many people had frequently put on a mask during COVID-19. It goes from level 1, Never, to level 7, Always. This will help aid my question on who was taking COVID-19 seriously.

===Distance Away Frequency===

```{r distance away frequency}
freq(as.ordered(data.xls$A31.6), plot = FALSE)

```

This table shows the frequency in which a person kept a distance of 1.5 meters away from others. The chart goes from level 1, Never, to level 7, Always. This will help me see who didn't wear their masks during COVID-19, which relates to my question of who didn't follow the guidelines provided to everyone across the globe.

=== Variable for Levels of Education===

```{r Education Level}
data.xls$LVL_EDU <- factor(ifelse(data.xls$A3.10 %in% c("1","2"), "High School and Below", "College and Above"))

summary(data.xls$LVL_EDU)
```

I am creating a variable named LVL_EDU to help me see who have educations from up to High School then College and higher.

===Variable for People of Only Religious Orientations===

```{r only religious}
data.xls$ONLY_R <- factor(ifelse(data.xls$A3.8 %in% c("1", "2","3","4","5","6","7","8","9","11"), "Has a Religious Belief", "No Beliefs"))

summary(data.xls$ONLY_R)
```

I am creating a variable named ONLY_R to group up who and who don't have a religious orientation.

===Variable for People Who Think COVID-19 was over exaggerated===

```{r}
data.xls$CVD_EXAG <- factor(ifelse(data.xls$`A5.3*` %in% c("1", "2","3","4","5"), "Reasonable Concern", "Out-of-Proportion"))

summary(data.xls$CVD_EXAG)
```

I am creating the variable CVD_EXAG to summarize the responses 1 = Not at all, 10= Extremely into two managed variables, options 1-5 being considered the people who believed that COVID-19 was not being over-exaggerated and options 6-10 being considered the people who thought the virus was being told to be a larger issue than it really was.

===Univariate Graphs===

```{r}
library(ggplot2)

data.xls %>%
  ggplot(aes(x = ONLY_R)) +
  geom_bar(fill = "forestgreen", color = "black") + 
  ggtitle("People Who Are of Religion") +
  xlab("Responses of Spiritual Belief") +
  ylab("Count of People") +
  theme_minimal() 

```

I am developing this graph to show how many people in the sample do and do not have any religious perspectives. I would like to correlate this graph later on with my other variables to see if there are any patterns between each other. One of my main concerns is that some people of religious backgrounds didn't think COVID-19 was a problem and that nothing would happen to each other. With this data gathered and graphed, I think it'll become easier to start comparing.

```{r}
library(ggplot2)

data.xls %>%
  ggplot(aes(x = LVL_EDU)) +
  geom_bar(fill = "darkred", color = "black") + 
  ggtitle("Levels of Education Within the Sample") +
  xlab("Different Levels of Education") +
  ylab("Count of People") +
  theme_minimal() 

```

This graph will help to visualize the differences between two major levels of education. The goal of this is to understand whether or not the levels education dictate what people believed during the COVID-19 pandemic and how serious the virus was. My graph here is just to show how many in the sample have received higher education or not.

```{r}
library(ggplot2)

data.xls %>%
  ggplot(aes(x = CVD_EXAG)) +
  geom_bar(fill = "forestgreen", color = "black") + 
  ggtitle("People Who Believed COVID-19 was Exagerated") +
  xlab("Opinions on COVID-19") +
  ylab("Count of People") +
  theme_minimal() 
```

This graph shows the many people who believed whether COVID-19 was overtly told as a massive issue and the many others who believed the concern of the virus was authentic. The main goal for this chart is to get an idea of the opinions many had about the virus when the issue was most prevalent.

====Bivariate Graph====

===Is Religion a Cause of Disregarding the COVID-19 Virus?===

```{r}
ggplot(data=data.xls, aes(x = factor(ONLY_R, labels = c("Not Religious", "Religious")), fill = factor(CVD_EXAG, labels = c("Exaggerated", "Not Exaggerated")))) +
    geom_bar(position = "dodge") +
    geom_text(stat='count', aes(label=..count..), position=position_dodge(0.9), vjust=-0.5) +
    labs(title = "Does Religion Impact the Perceptions of COVID-19",
         x = "Religious Beliefs",
         y = "Count of People",
         fill = "COVID-19 View") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1") +
    theme(panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "black", fill = NA, size = 1))
```

Through my findings I have found out that regardless of religion, a majority within a sample have agreed that COVID-19 was as serious as it was. I wanted to make this graph because through my own experiences during the pandemic, I witnessed many people disregarding the virus and explaining that their place of worship denied the virus being real or even an issue. This graph helps me see that it isn't really like this everywhere else.

===Did the Level of Education Affect the Perception of COVID-19 as a serious issue?===

```{r}
ggplot(data=data.xls, aes(x = factor(LVL_EDU, labels = c("College and Higher", "High School and Lower")), fill = factor(CVD_EXAG, labels = c("Exaggerated", "Not Exaggerated")))) +
    geom_bar(position = "dodge") +
    geom_text(stat='count', aes(label=..count..), position=position_dodge(0.9), vjust=-0.5) +
    labs(title = "Does Level of Education Impact How COVID-19 was perceived?",
         x = "Levels of Education",
         y = "Count",
         fill = "COVID-19 View") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1") +
    theme(panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "black", fill = NA, size = 1))
```

A common issue that's been witnessed among many is how a lack of education can lead people to be ignorant about certain theories, ideas or health issues. One of the most leading issues would be people who practice anti-vaxxing, the act of refusing to use a vaccine for medical problems. The cause of this is usually the lack of education to help a person understand why they should be taking these medical solutions. This is one of the biggest barriers in vaccine uptake. I wanted to see if the differences in education within the sample saw COVID-19 as an issue or not. Through my research and the data provided we can see that many, highly educated or not, did see the virus as an issue to take seriously.
