---
title: "Working with Data"
format: 
  html:
    self-contained: true
  pdf: 
    number-sections: true
    toc:
    pdf-engine: xelatex
editor: visual
---

```{r library}
library(tidyverse)
library(descr)
library(knitr)
library(dplyr)
library(readr)
library(readxl)
```

```{r load-data}
file_path <- "/cloud/project/data/data.xls"

data.xls <- read_excel(file_path)

head(data.xls)
```

```{r religious orientations}
freq(as.ordered(data.xls$A3.8), plot = FALSE)
```

\^This table explains the frequency of different religious orientations among a sample.

```{r education levels}
freq(as.ordered(data.xls$A3.10), plot = FALSE)

```

\^This table explains how much education a person within a sample has achieved.

```{r COVID-19 anxiety}
#
freq(as.ordered(data.xls$`A5.3*`), plot = FALSE)

```

\^This table explains the levels of anxiety a person experienced from the COVID-19 pandemic.

```{r mask frequency}
freq(as.ordered(data.xls$A31.4), plot = FALSE)

```

\^This table shows the frequency in which a person used a mask during the COVID-19 pandemic.

```{r distance away frequency}
freq(as.ordered(data.xls$A31.6), plot = FALSE)

```

\^This table shows the frequency in which a person kept a distance of 1.5 meters away from others.

```{r high school and below}
data.xls$HS_NBELOW <- factor(ifelse(data.xls$A3.10 %in% c("1", "2"), "Yes", "No"))

summary(data.xls$HS_NBELOW)
```

```{r colleg and above}
data.xls$SOME_CLGNABOVE <- factor(ifelse(data.xls$A3.10 %in% c("3", "4", "5", "6", "7", "8"), "Yes", "No"))

summary(data.xls$SOME_CLGNABOVE)
```

```{r religious excluding}
data.xls$religious <- factor(ifelse(data.xls$A3.8 %in% c("1", "2","3","4","5","6","7","8","9","11"), "Yes", "No"))
summary(data.xls$religious)
```
