---
title: "ANOVA"
output: learnr::tutorial
runtime: shiny_prerendered
tutorial:
  id: anova_tutorial
  title: "ANOVA Using Your Own Data"
  description: "Learn how to run an ANOVA and check assumptions using R"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
library(learnr)
library(ggplot2)
library(gplots)
library(sciplot)
library(car)
library(lsr)
```

```{r include=FALSE}
# Set seed for reproducibility
set.seed(123)

# Create factors
drug <- factor(rep(c("placebo", "anxifree", "joyzepam"), each = 20))
therapy <- factor(rep(rep(c("no.therapy", "CBT"), each = 10), times = 3))

# Generate mood gain scores with some real differences:
# Let's assume:
# - joyzepam > anxifree > placebo
# - CBT is generally more effective than no therapy
# - Slight interaction: anxifree + CBT gives a bigger-than-expected boost

mood.gain <- c(
  rnorm(10, mean = 0.2, sd = 0.3),  # placebo, no.therapy
  rnorm(10, mean = 0.5, sd = 0.3),  # placebo, CBT
  rnorm(10, mean = 0.7, sd = 0.3),  # anxifree, no.therapy
  rnorm(10, mean = 1.2, sd = 0.3),  # anxifree, CBT (interaction boost)
  rnorm(10, mean = 1.5, sd = 0.3),  # joyzepam, no.therapy
  rnorm(10, mean = 1.6, sd = 0.3)   # joyzepam, CBT
)

# Combine into a data frame
clin.trial <- data.frame(
  drug = drug,
  therapy = therapy,
  mood.gain = mood.gain
)

# Preview the data
head(clin.trial)
```

## Why Use ANOVA Instead of Multiple t-Tests?

When comparing more than two group means, it's tempting to just run a
bunch of t-tests — for example, comparing Group 1 vs Group 2, Group 1 vs
Group 3, and so on. But this approach has a serious drawback:

**Each t-test carries a risk of a Type I error (false positive). The
more tests you run, the greater your overall chance of making at least
one mistake.**

Let’s look at an example:

-    Imagine you're comparing 3 groups.

-   To compare them using t-tests, you’d need 3 comparisons:

    -    Group 1 vs Group 2

    -    Group 1 vs Group 3

    -    Group 2 vs Group 3

Each test typically uses a significance level of 0.05, meaning there’s a
5% chance of finding a significant result by chance — even if there's no
real difference.

**When you run 3 tests, the chance of at least one false positive is
greater than 5%.**

As you add more groups (and more tests), this problem gets worse — this
is known as the **inflation of the familywise error rate**.

### How ANOVA Solves the Problem

**ANOVA (Analysis of Variance)** solves this by testing **all groups at
once** using a single overall test:

-    It compares **variance between group means** to **variance within
    groups**

-   It produces a single **F-test** to tell us whether *any* group means
    are significantly different

-   If the F-test is significant, we can follow up with **post-hoc
    tests** to see *which* groups differ, while controlling for error

## ANOVAs

### Introduction to ANOVA

In statistics, when we want to compare **more than two group means**, we
use **Analysis of Variance (ANOVA)**. While a t-test can compare two
groups, running multiple t-tests inflates the Type I error rate. ANOVA
allows us to test all groups simultaneously while controlling for error.

-   **ANOVA compares the variance *between groups* to the variance
    *within groups*.**

-   We use the **F-ratio** to evaluate whether observed group
    differences are larger than we'd expect due to random chance.

#### Hypotheses for a One-Way ANOVA:

-    **Null hypothesis (H₀):** All group means are equal.\
    μ1​=μ2​=μ3

-   **Alternative hypothesis (H₁):** At least one group mean is
    different.\
    (Not all μs are equal, but doesn't specify which ones.)

This sets the stage for using statistical modeling to determine whether
group membership explains a meaningful portion of the variability in the
outcome.

### Descriptive Statistics and Visuals

Before running ANOVA, it’s important to **explore the data**.

***Descriptive Tools***

-   xtabs(): Cross-tabulation to check how many observations are in each
    group.

-   aggregate(): Calculates group means (or other summaries like SD).

-   plotmeans() from the gplots package: Plots means with confidence
    intervals.

```{r example-describe-ANOVA}
# Check number of cases in each group
xtabs(~drug, data = clin.trial)
```

```{r}
# Calculate group means
aggregate(mood.gain ~ drug, data = clin.trial, mean)
```

```{r}
# Plot means with confidence intervals
plotmeans(mood.gain ~ drug, data = clin.trial)
```

These steps help verify:

-   **Balanced groups** (equal sample sizes)

-   **Group-level trends** before inference

### Model Calculations

What's happening with ANOVA calculations? These are the components:

-   **Sum of Squares Between (SSB):** Variability due to differences
    between group means

-   **Sum of Squares Within (SSW):** Residual/error variability

-   **Mean Squares (MS):** SS divided by degrees of freedom

-   **F-statistic:** Ratio of MS between to MS within

```{r example-ANOVA-calculations}
model <- aov(mood.gain ~ drug, data = clin.trial)
summary(model)
```

This gives us:

-   Sum of Squares (Between and Within)

-   Mean Squares

-   F-statistic and *p*-value

### Post-Hoc Analysis

If the overall ANOVA is significant, we **don't yet know which groups
differ**. That's where post-hoc tests come in.

**Only** if the ANOVA is significant would we follow up with comparisons
between groups. Post-hoc comparison would be justified because the
overall ANOVA only tells us that **some group means differ**—not which
ones.

-   Use posthocPairwiseT() from the lsr package.

-   It runs **pairwise comparisons** between groups while adjusting for
    multiple testing.

This function performs multiple pairwise comparisons while adjusting for
Type I error. It tells us *which* groups differ.

### Assumption Checks

Before reporting results, we **check assumptions** and **compute effect
size**:

***Effect Size:***

-   etaSquared(model) reports **η²**, the proportion of variance in the
    outcome explained by group membership.

-   Interpreted like R² in regression (e.g., η² = 0.71 = 71% variance
    explained).

***Assumption Checks:***

-   leveneTest() from the car package tests **homogeneity of variance**.

-   shapiro.test() tests **normality of residuals** (assuming
    independence).

### Template Code

```{r example-ANOVA-code, eval = FALSE}
# Load necessary libraries
library(lsr)       # for etaSquared, posthocPairwiseT
library(gplots)    # for plotmeans
library(car)       # for Levene's test

# One-way ANOVA: comparing means of one outcome across one grouping variable

# View number of cases per group
xtabs(~ grouping_variable, data = your_data)

# Get mean of outcome variable for each group
aggregate(outcome_variable ~ grouping_variable, data = your_data, mean)

# Get standard deviation of outcome variable for each group
aggregate(outcome_variable ~ grouping_variable, data = your_data, sd)

# Plot group means with 95% confidence intervals
plotmeans(outcome_variable ~ grouping_variable, 
          data = your_data,
          xlab = "Group Name (e.g., Treatment Type)",
          ylab = "Outcome Variable (e.g., Mood Gain)",
          main = "Outcome by Group with 95% CI")

# Run the one-way ANOVA
model <- aov(outcome_variable ~ grouping_variable, data = your_data)

# Summary of ANOVA table
summary(model)

# Calculate eta squared effect size
etaSquared(model)

# Post-hoc comparisons (pairwise t-tests)
posthocPairwiseT(model)

# Levene's test for equal variances
leveneTest(outcome_variable ~ grouping_variable, data = your_data)

# Shapiro-Wilk test for normality of residuals
shapiro.test(residuals(model))
```

------------------------------------------------------------------------

## Factorial ANOVA

### Introduction to Factorial ANOVA

**Factorial ANOVA** allows us to study:

-   **Two or more categorical IVs (factors)**

-   **Main effects** of each IV

-   **Interaction effect** between IVs

Example:

-   drug (placebo, anxifree, joyzepam)

-   therapy (no.therapy, CBT)

The design is described as a **3×2 factorial ANOVA**, examining whether
drug and therapy affect mood.gain, alone, or interactively.

### Model Calculations

You can include main effects and interactions like this:

```{r example-Factoral-model}
model2 <- aov(mood.gain ~ drug * therapy, data = clin.trial)
summary(model2)
```

You can use either drug + therapy + drug:therapy or simply drug \*
therapy in the formula.

This shows:

-   The **main effects** of drug and therapy appear separately.

-   The **interaction term** shows whether the effect of one variable
    depends on the level of the other.

### Interaction Plots

Plotting helps visually assess interaction effects:

```{r example-Factoral-plot}
library(sciplot)
lineplot.CI(x.factor = clin.trial$drug,
            group = clin.trial$therapy,
            response = clin.trial$mood.gain)
```

-   If the lines are not parallel, that suggests an interaction between
    the two variables.

### Model Interpretation

To extract model-predicted **group means and confidence intervals**:

```{r example-Factoral-means}
library(effects)
eff <- effect("drug*therapy", model2)
summary(eff)
```

-   These are **estimated marginal means**, accounting for all terms in
    the model.

-   Useful for **reporting group differences with precision**.

effect() is useful especially when group sizes are unequal or when
interactions exist.

### Assumption Checks

Same as one-way ANOVA, we need to test assumptions:

```{r example-Factoral-assumption}
library(car)
leveneTest(mood.gain ~ drug * therapy, data = clin.trial)
shapiro.test(residuals(model2))
```

**Ignoring assumptions** like homogeneity can inflate Type I errors.
These tests confirm that the model’s assumptions are not violated.

### Effect Sizes

You can quantify how much each effect (main or interaction) contributes
to outcome variability:

```{r example-Factoral-effect-size}
etaSquared(model2)
```

-   Outputs **η²** and **partial η²**.

-   Partial η² is especially useful in **factorial designs**, where you
    isolate each term’s effect *controlling for others*.

This provides both **eta squared** and **partial eta squared**, which
show the proportion of variance explained by each effect. Use this to
determine if an effect is not only significant, but meaningful.

### Template Code

```{r example-Factorial-code, eval = FALSE}
# Load libraries
library(sciplot)   # for lineplot.CI
library(effects)   # for estimated marginal means

# Factorial ANOVA: comparing effects of two categorical variables and their interaction

# View number of cases per combination of the two grouping variables
xtabs(~ factor_A + factor_B, data = your_data)

# Means for each combination of factor levels
aggregate(outcome_variable ~ factor_A + factor_B, data = your_data, mean)

# Run the factorial ANOVA
factorial_model <- aov(outcome_variable ~ factor_A * factor_B, data = your_data)

# Summary of model (includes main effects and interaction)
summary(factorial_model)

# Plot interaction with 95% CI error bars
lineplot.CI(x.factor = your_data$factor_A,
            group = your_data$factor_B,
            response = your_data$outcome_variable,
            xlab = "Factor A (e.g., Treatment Type)",
            ylab = "Outcome Variable (e.g., Performance)",
            main = "Interaction Plot")

# Get eta squared effect sizes for all terms
etaSquared(factorial_model)

# Estimated marginal means and confidence intervals for each group
effect_model <- effect("factor_A*factor_B", factorial_model)
summary(effect_model)

# Levene’s test for homogeneity of variance across all groups
leveneTest(outcome_variable ~ factor_A * factor_B, data = your_data)

# Normality check on residuals
shapiro.test(residuals(factorial_model))
```
